Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-26 write-ups/FSE/paper.tex
Shortenings.

\newcommand{\myparagraph}[1]{\paragraph{#1.}}
%{\noindent{\bf\emph{#1}}\hspace{1em}}

@@

%\vspace{-1mm}
\myparagraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms
demand fair scheduling of workgroups, but for good reasons
GPU vendors will not commit even to the guarantees of the
occupancy-bound execution model.

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A kernel
that requires fair scheduling is identified as \emph{cooperative}, and written using two additional
language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates that the workgroup
would ideally continue executing, but that the
scheduler may preempt the workgroup; the cooperative kernel
must be prepared to deal with either scenario.
%
At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$ to indicates that the
kernel is prepared to proceed with the existing set of workgroups, but
is able to benefit from one or more additional workgroups
commencing execution directly after the $\offerfork$ program point.

The use of $\offerfork$ and $\offerkill$ creates a contract between
the scheduler and the cooperative kernel.  Functionally, the scheduler
must guarantee that the workgroups executing a cooperative kernel are
fairly scheduled, while the cooperative kernel must be robust to
workgroups leaving and joining the computation in response to
$\offerkill$ and $\offerfork$.  Non-functionally, a cooperative kernel
must ensure that $\offerkill$ is executed frequently enough such that
the scheduler can accommodate soft-real time constraints,
e.g.\ allowing a smooth frame-rate for graphics, or allowing compute
units to be powered down quickly when required.  The scheduler should
respond to $\offerkill$ and $\offerfork$ calls in a manner that avoids
under-utilisation of hardware resources by the cooperative kernel,
e.g.\ the scheduler should not kill workgroups unless they are
required for another purpose, and should facilitate forking of
additional workgroups when possible.

\myfiglong{\ref{fig:overview}} shows an example execution of how the
GPU compute units might be shared between a cooperative kernel and a
graphics task. Notice that workgroups executing the cooperative kernel
are killed at a $\offerkill$ command to make room for a graphics
task. After the graphics task, the previously killed workgroups are
restored to executing the cooperative kernel when a workgroup
executing the cooperative kernel calls $\offerfork$. The order in
which workgroups are killed is important and discussed in
\mysec{\ref{sec:semantics}}. The \emph{gather time} is the time
between when resources are requested and when the application has
surrendered the resources via $\offerkill$. To satisfy soft-real time
constraints, this time should be low; our experimental study
(\mysec{\ref{sec:responsiveness}}) shows that in practice, the
gather-time for our applications is acceptable for a range of graphics
workloads.

The cooperative kernels programming model has several appealing
properties:

@@

\item The model has no impact on the development of regular
  (non-cooperative) compute and graphics kernels: these can be programmed exactly as they
  are now.
%, as can graphics shading routines written using APIs such as
%OpenGL, Vulkan and DirectX.

@@

\item Cooperative kernels can be implemented over the occupancy\-/bound
  execution model that current GPUs provide: our prototype implementation requires
  no additional hardware or driver support.

\item If hardware support for preemption is available it can be leveraged to implement cooperative kernels efficiently, and cooperative kenrnels can avoid the overhead of unnecessary preemptions by allowing the programmer to communicate ``smart'' preemption points.

\item Our experiments with a range of applications \cutthree{using three GPUs} show that the model can enable efficient multitasking of cooperative and non-cooperative
  tasks.

@@

Placing the new primitives requires manual effort, but our experience porting a representative set of
GPU-accelerated irregular algorithms to use cooperative kernels
(\mysec\ref{sec:portingalgorithms}) suggests that this is straightforward in practice.

In summary, our main contributions are:
\emph{cooperative kernels}, an extended GPU programming model that supports the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}); a \emph{prototype implementation} of cooperative
  kernels on top OpenCL 2.0
  (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms \cutthree{across three GPUs} (\mysec\ref{sec:experiments}).

>>>>

%\newcommand{\myparagraph}[1]{\paragraph{#1.}}
\newcommand{\myparagraph}[1]{\vspace{1mm}\noindent\emph{\textbf{#1.}}}
%{\noindent{\bf\emph{#1}}\hspace{1em}}

@@

%\vspace{-1mm}
\myparagraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms
demand fair scheduling,
% of workgroups,
but for good reasons
GPU vendors will not commit even to the guarantees of the
occupancy-bound execution model.
%
We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.

A kernel
that requires fair scheduling is identified as \emph{cooperative}, and written using two additional
language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.
%
Where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates that the workgroup
would ideally continue executing, but that the
scheduler may preempt the workgroup; the cooperative kernel
must be prepared to deal with either scenario.
%
Where the cooperative kernel could use additional resources, a workgroup can execute
$\offerfork$ to indicates that the
kernel is prepared to proceed with the existing set of workgroups, but
is able to benefit from one or more additional workgroups
commencing execution directly after the $\offerfork$ program point.

The use of $\offerfork$ and $\offerkill$ creates a contract between
the scheduler and the cooperative kernel.  Functionally, the scheduler
must guarantee that the workgroups executing a cooperative kernel are
fairly scheduled, while the cooperative kernel must be robust to
workgroups leaving and joining the computation in response to
$\offerkill$ and $\offerfork$.  Non-functionally, a cooperative kernel
must ensure that $\offerkill$ is executed frequently enough such that
the scheduler can accommodate soft-real time constraints,
e.g.\ allowing a smooth frame-rate for graphics.
%, or allowing compute
%units to be powered down quickly when required.
In return, the scheduler should allow the cooperative kernel to utilise
hardware resources where possible, killing workgroups only when demanded by other tasks, and forking additional workgroups when possible.

\myfiglong{\ref{fig:overview}} illustrates sharing of GPU compute units between a cooperative kernel and a
graphics task. Workgroups 2 and 3 of the cooperative kernel
are killed at an $\offerkill$ to make room for a graphics
task. The workgroups are subsequently restored to the cooperative kernel when workgroup 0 calls $\offerfork$. The order in
which workgroups are killed is important and discussed in
\mysec{\ref{sec:semantics}}. The \emph{gather time} is the time
between resources being requested and the application surrendering them via $\offerkill$. To satisfy soft-real time
constraints, this time should be low; our experimental study
(\mysec{\ref{sec:responsiveness}}) shows that in practice, the
gather-time for our applications is acceptable for a range of graphics
workloads.

The cooperative kernels model has several appealing
properties:

@@

\item The model has no impact on the development of regular
  (non-cooperative) compute and graphics kernels.
%, as can graphics shading routines written using APIs such as
%OpenGL, Vulkan and DirectX.

@@

\item Cooperative kernels can be implemented over the occupancy\-/bound
  execution model provided by current GPUs: our prototype implementation uses no special hardware/driver support.

\item If hardware support for preemption \emph{is} available, it can be leveraged to implement cooperative kernels efficiently, and cooperative kernels can avoid unnecessary preemptions by allowing the programmer to communicate ``smart'' preemption points.

@@

Placing the new primitives requires manual effort, but is straightforward for the representative set of
GPU-accelerated irregular algorithms that we have ported to use cooperative kernels.  Our experiments show that the model can enable efficient multitasking of cooperative and non-cooperative tasks.

In summary, our main contributions are:
\emph{cooperative kernels}, an extended GPU programming model that supports the scheduling requirements of blocking algorithms (\mysec\ref{sec:cooperativekernels}); a \emph{prototype implementation} of cooperative
  kernels on top OpenCL 2.0
  (\mysec\ref{sec:implementation}); and \emph{experiments} assessing the overhead and responsiveness of the cooperative kernels approach over a set of irregular algorithms \cutthree{across three GPUs} (\mysec\ref{sec:experiments}), including a best-effort comparison with the efficiency afforded by hardware-supported preemption available on Nvidia GPUs.
